---
title: "DS Project"
author: "Lucas Rapeepat, Ian Stripe"
format: 
  html:
    toc: true
    embed-resources: true
    code-fold: true
    code-line-numbers: true
    code-summary: "SHOW ME THE CODE!"
    code-tools: true
    theme: yeti
editor: visual
---

# Introduction

In the contemporary landscape of American family life, the cost of childcare emerges as a pivotal factor influencing economic stability, parental employment, and, ultimately, child welfare. Access to affordable, quality childcare has been linked to a range of positive outcomes, including increased workforce participation, particularly among women, and enhanced child development. Yet, for many families across the United States, childcare remains a significant financial burden, absorbing a disproportionate share of household income. This project seeks to illuminate the complexities surrounding childcare costs, shedding light on their variability across different geographical regions and over time.

This analysis is anchored in data drawn from the National Database of Childcare Prices, the most comprehensive federal repository of childcare price information at the county level. Spanning from 2008 to 2018, this database offers a granular look at childcare costs across various dimensions: by provider type, the age of children, and county characteristics. The project utilizes two primary data sets: childcare_costs.csv, containing detailed childcare price data, and counties.csv, which provides a suite of county-level attributes. These datasets are unified through the county_fips_code, enabling a holistic examination of how childcare costs correlate with a myriad of socioeconomic factors.

The goal of this project is not merely to chart the ebb and flow of childcare expenses but to understand the broader implications of these costs on families and communities. By employing a robust analytical framework, this study will explore several key questions: How have childcare costs evolved over the last decade? Do these costs exhibit significant variability across states or regions? What socioeconomic variables—such as unemployment rates, median household income, or racial composition—interact most dynamically with childcare expenses?

The answers to these questions promise not only to advance our understanding of childcare economics but also to inform policy discussions aimed at making quality childcare more accessible and affordable. Through a meticulous analysis of available data, this project endeavors to contribute a meaningful voice to the ongoing dialogue about how best to support American families in balancing the demands of work and child-rearing.














# Step 2

```{r}
#Set-up
library(tidyverse)
counties <- read.csv("C:/Users/Lucas Rapeepat/Downloads/counties.csv")
costs <- read.csv("C:/Users/Lucas Rapeepat/Downloads/childcare_costs.csv")


#Use drop_na to pre-process data
```



Data Pre-Processing

The foundational step in our analysis involved the meticulous preparation of two pivotal datasets, each serving as a cornerstone for our exploration of childcare costs across the United States. The first dataset, childcare_costs.csv, encompasses a comprehensive array of childcare pricing information, delineated by provider type, child age, and the specific characteristics of each county. This dataset spans from 2008 to 2018, offering a decade's worth of insights into the evolution of childcare expenses. With over 30,000 rows, it encapsulates yearly data for numerous counties, presenting an opportunity to dissect the financial aspects of childcare with unprecedented granularity.

Our second dataset, counties.csv, complements the first by providing a rich tapestry of county-level attributes. This includes, but is not limited to, demographic breakdowns, economic indicators, and social characteristics. The fusion of these datasets via the county_fips_code, a unique identifier for each county, allows for a nuanced analysis that transcends mere cost examination to include socio-economic factors.

Joining Data Sets
The initial phase of data pre-processing involved the unification of the childcare_costs.csv and counties.csv datasets. This was accomplished through an inner join operation, ensuring that only records with matching county_fips_code in both datasets were retained. This meticulous merging process was instrumental in creating a cohesive dataset that aligns childcare cost data with corresponding county characteristics, laying the groundwork for a comprehensive analysis.

Removing Missing Data
A critical step in data cleaning was the systematic removal of missing data. Given the expansive nature of our datasets, missing values were inevitable. However, to ensure the integrity and reliability of our analysis, rows containing missing values—particularly in key variables of interest such as childcare costs and county characteristics—were excluded. This pruning was essential for maintaining a high standard of accuracy in our subsequent analyses.

Filtering Data
To tailor our dataset to the specific scope of our study, we applied filtering criteria, honing in on the most relevant records. This involved selecting data pertinent to our analysis timeframe, from 2008 to 2018, and focusing on variables critical to our research questions. Such filtering was pivotal in refining our dataset, ensuring that our analysis remained focused and relevant.


## Joining the Tables and filtering out N/A

```{r}
ccosts <- costs |>
  inner_join(counties, by = "county_fips_code") |>
  drop_na()

```


# Section 3

Ideas: 










